---
title: "Mohamad_Nayal_HW7"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), warning = FALSE, message = FALSE, tidy=TRUE) 
```
# Introduction
 
 
# Analysis

## About the Data
```{r echo=FALSE}
#Importing packages
library(dplyr)
library(caret)
library(rpart.plot)
library(ggplot2)
library(e1071)
library(FactoMineR)
library(plotly)
library(rpart)
library(rattle)
library(ggplot2)
library(randomForest)
```

```{r echo =FALSE}
library(parallel)
library(iterators)
library(foreach)
library(doParallel)

split = detectCores()
eachStart = 25

cl = makeCluster(split)
init = clusterEvalQ(cl, { library(MASS); NULL })
```


```{r echo=FALSE}
#importing dataset
diabetes_complete <- read.csv("diabetes_012_health_indicators_BRFSS2015.csv")
cat("Number of rows in dataset: ",nrow(diabetes_complete), "\n")
diabetes_complete <- na.omit(diabetes_complete)
cat("Number of rows after removing NAs: ", nrow(diabetes_complete))
```

```{r echo=FALSE}
#getting number of healthy
cat("Number of healthy individuats: ", nrow(diabetes_complete[diabetes_complete$Diabetes_012==0,]), "\n")

#getting number of pre-diabetics
cat("Number of Pre-Diabetic individuats: ",nrow(diabetes_complete[diabetes_complete$Diabetes_012==1,]), "\n")

#getting number of diabetics
cat("Number of Diabetic individuats: ", nrow(diabetes_complete[diabetes_complete$Diabetes_012==2,]))
```

```{r}
head(diabetes_complete)
```

```{r}
#create combined balanced dataset
#get all healthy individuals
healthy <- diabetes_complete[diabetes_complete$Diabetes_012==0,]
#sample 4630 healthy individuals by getting range from 1 to length of healthy, sampling  and then using them as #index
healthy <- healthy[sample(1:nrow(healthy), 4630, replace = FALSE),]

#repeat for prediab
preDiab <- diabetes_complete[diabetes_complete$Diabetes_012==1,]
preDiab <- preDiab[sample(1:nrow(preDiab), 4630, replace = FALSE),]

#repeat for Diabetics
diabetics<- diabetes_complete[diabetes_complete$Diabetes_012==2,]
diabetics <- diabetics[sample(1:nrow(diabetics), 4630, replace = FALSE),]

#combine to form balanced dataframe
balanced <- rbind(healthy, preDiab)
balanced <- rbind(balanced, diabetics)

head(balanced)

```

```{r echo =FALSE}
#printing balanced numbers 
cat("Number of healthy individuats: ",nrow(balanced[balanced$Diabetes_012==0,]), "\n")
cat("Number of Pre-Diabetic individuats: ",nrow(balanced[balanced$Diabetes_012==1,]), "\n")
cat("Number of Diabetic individuats: ",nrow(balanced[balanced$Diabetes_012==2,]), "\n")
```

```{r}

diabetes_complete

balanced$Diabetes_012 <- as.factor(balanced$Diabetes_012)
balanced$HighBP <- as.factor(balanced$HighBP)
balanced$HighChol <- as.factor(balanced$HighChol)
balanced$CholCheck <- as.factor(balanced$CholCheck)
balanced$Smoker <- as.factor(balanced$Smoker)
balanced$Stroke <- as.factor(balanced$Stroke)
balanced$HeartDiseaseorAttack <- as.factor(balanced$HeartDiseaseorAttack)
balanced$PhysActivity <- as.factor(balanced$PhysActivity)
balanced$Fruits <- as.factor(balanced$Fruits)
balanced$Veggies <- as.factor(balanced$Veggies)
balanced$HvyAlcoholConsump <- as.factor(balanced$HvyAlcoholConsump)
balanced$AnyHealthcare <- as.factor(balanced$AnyHealthcare)
balanced$NoDocbcCost <- as.factor(balanced$NoDocbcCost)
balanced$GenHlth <- as.factor(balanced$GenHlth)
balanced$DiffWalk <- as.factor(balanced$DiffWalk)
balanced$Sex <- as.factor(balanced$Sex)
balanced$Age <- as.factor(balanced$Age)
balanced$Education <- as.factor(balanced$Education)
balanced$MentHlth <- as.factor(balanced$MentHlth)
balanced$PhysHlth <- as.factor(balanced$PhysHlth)
balanced$Income <- as.factor(balanced$Income)
```

```{r}
str(balanced)
```

``````{r echo=FALSE}

balanced$GenHlth <- plyr::mapvalues(balanced$GenHlth, from = c("1","2","3","4","5"), to = c("Excellent", "Very Good", "Good", "Fair", "Poor"))
GenHlthPlot <- ggplot(balanced, aes(x=GenHlth)) + 
    geom_bar(fill = "red") +
    ggtitle("General Health")+
    ylab("Number of People")+theme(plot.title = element_text(hjust = 0.5))
GenHlthPlot
````
```{r echo=FALSE}

bmiPlot <- ggplot(balanced, aes(x=BMI)) + 
    geom_bar(fill = "blue") +
    ggtitle("BMI")+
    xlab("Body Mass Index")
    ylab("Number of People")+theme(plot.title = element_text(hjust = 0.5))
    
bmiPlot
```

```{r}
#Making BMI Discrete USing CDC BMI categories
#https://www.cdc.gov/healthyweight/assessing/bmi/adult_bmi/index.html
balanced$BMI <- cut(balanced$BMI, breaks = c(10,18.5,24.9,29.9,34.9,39.9,40), 
                labels=c("Underweight", "Normal weight", "Pre-obesity", "Obesity class I", "Obesity class II","Obesity class III"))

#bank$age <- cut(bank$age, breaks = c(10,20,30,40,50,60,70), 
 #               labels=c("teens", "twenties", "thirties", "fourties", "fifties", 
  #                       "sixties"))
balanced <- na.omit(balanced)
```


```{r echo=FALSE}
bmiPlot2 <- ggplot(balanced, aes(x=BMI)) + 
    geom_bar(fill = "blue") +
    ggtitle("BMI")+
    xlab("Body Mass Index")
    ylab("Number of People")+theme(plot.title = element_text(hjust = 0.5))
    
bmiPlot2
```

```{r}
balanced <- subset(balanced, BMI != "Underweight")
balanced <- subset(balanced, BMI != "Obesity class III")
balanced$BMI <- droplevels(balanced$BMI)
```

```{r echo=FALSE}
bmiPlot3 <- ggplot(balanced, aes(x=BMI)) + 
    geom_bar(fill = "blue") +
    ggtitle("BMI")+
    xlab("Body Mass Index")
    ylab("Number of People")+theme(plot.title = element_text(hjust = 0.5))
    
bmiPlot3
```

```{r}
balanced <- na.omit(balanced)
```

```{r}
summary(balanced$BMI)
```


```{r}
MenHlthPlot <- ggplot(balanced, aes(x=MentHlth)) + 
    geom_bar(fill = "blue") +
    ggtitle("Mental Health")+
    xlab("Number of days of peeor mental health last month")
    ylab("Number of People")+
    theme(plot.title = element_text(hjust = 0.5))
    
MenHlthPlot
```
```{r}
balanced$MentHlth <- as.numeric(balanced$MentHlth)
balanced$MentHlth <- cut(balanced$MentHlth, breaks = c(0,2,5,10,15,20,25, 31), 
                labels=c("0-2", "2-5", "5-10", "`0-15", "15-20","20-25", "25-31"))

MenHlthPlot2 <- ggplot(balanced, aes(x=MentHlth)) + 
    geom_bar(fill = "blue") +
    ggtitle("Mental Health")+
    xlab("Number of days of peeor mental health last month")
    ylab("Number of People")+
    theme(plot.title = element_text(hjust = 0.5))
    
MenHlthPlot2
```



```{r  echo=FALSE}
balanced$Income <- plyr::mapvalues(balanced$Income, from = c("1","2","3","4","5","6","7","8"), to = c("1 - less than 10K", "2", "3", "4", "5- less than 35K", "6", "7", "8 more than 75k"))

#Example code for creating waffle chart obtained from 
#<https://www.geeksforgeeks.org/r-waffle-chart/>

waffle::waffle(table(balanced$Income)/100, rows=10, size=1, 
       colors=c("#44D2AC", "#E48B8B", "#B67093", 
                "#3A9ABD", "#CFE252", "cadetblue1", "darkgoldenrod1", "darkorchid2"), 
       title="Income Groups bifurcation", 
       xlab="1 square = 100 people")
```

```{r echo=FALSE}
balanced$Sex <- plyr::mapvalues(balanced$Sex, from = c("0","1"), to = c("Female", "Male"))
sexPlot <- ggplot(balanced, aes(x="", y=Sex, fill=Sex)) + 
    geom_bar(stat="identity", width=1)+
    coord_polar("y", start=0)+
    ggtitle("Sex of Participants")+
    theme(plot.title = element_text(hjust = 0.5))+
    theme_void()
    
sexPlot

```


```{r}

#running a t test to see if difference in BMI based off those who eat veggies

t.test(diabetes_complete$BMI[diabetes_complete$Veggies==0],diabetes_complete$BMI[diabetes_complete$Veggies==1])

```





```{r}

noveggielist <- diabetes_complete$BMI[diabetes_complete$Veggies==0]
veggielist <- diabetes_complete$BMI[diabetes_complete$Veggies==1]

par(mfrow=c(1,1)
boxplot(noveggielist,veggielist, main = "BMI of veggie eater vs non veggie eater", ylab = "BMI",names = c("Non-Veggie eaters", "Veggie eaters"))

```

```{r}

bmiVeg <- ggplot(balanced,aes(x=BMI, fill=Veggies))+
    geom_bar(position="dodge")+
    ggtitle("BMI vs Vegetable Diet")+
    ylab("Number of People")+
    theme(plot.title = element_text(hjust = 0.5))+
    scale_fill_manual(name="Veggie",values=c("red", "green"),labels=c("No Veggies", "Veggie Eater"))
bmiVeg
```



```{r}
#counts of diet preference
nrow(diabetes_complete[diabetes_complete$Veggies==0,])
nrow(diabetes_complete[diabetes_complete$Veggies==1,])
```
```{r}

#running a t test to see if difference in BMI based off those who eat veggies

t.test(diabetes_complete$BMI[diabetes_complete$Fruits==0],diabetes_complete$BMI[diabetes_complete$Fruits==1])

```

```{r}

nofruitlist <- diabetes_complete$BMI[diabetes_complete$Fruits==0]
fruitlist <- diabetes_complete$BMI[diabetes_complete$Fruits==1]

par(mfrow=c(1,1))
boxplot(nofruitlist,fruitlist, main = "BMI of fruit eaters vs non-fruit eaters", ylab = "BMI",names = c("Non-Fruit eaters", "Fruit eaters"))

```

```{r echo=FALSE}
bmiFruit <- ggplot(balanced,aes(x=BMI, fill=Fruits))+
    geom_bar(position="dodge")+
    ggtitle("BMI vs Fruits Diet")+
    ylab("Number of People")+
    theme(plot.title = element_text(hjust = 0.5))+
    scale_fill_manual(name="Veggie",values=c("black", "pink"),labels=c("No Fruit", "Fruits in Diet"))
bmiFruit
```
```{r echo=FALSE}

diabetesAge <- ggplot(balanced,aes(x=Age, fill=Diabetes_012)) +
    geom_bar()+
    ggtitle("Age vs Number of Diabetics")+
    ylab("Number of People")+
    theme(plot.title = element_text(hjust = 0.5))
diabetesAge

diabetesBMI <- ggplot(balanced,aes(x=BMI, fill=Diabetes_012)) +
    geom_bar(position="dodge")+
    ggtitle("BMI vs Number of Diabetics")+
    ylab("Number of People")+
    theme(plot.title = element_text(hjust = 0.5))
diabetesBMI
```




```{r echo=FALSE}
#sampling balanced dataset to take out 66% for training 
smp_size <- floor(0.66 * nrow(balanced))

## set the seed to make your partition reproducible
set.seed(123)
train_ind <- sample(seq_len(nrow(balanced)), size = smp_size)

train <- balanced[train_ind, ]
test <- balanced[-train_ind, ]
train2 <- subset(train,select=-c(GenHlth))

# Creating a control with cross validation of 3
control <- trainControl(method ='cv',number = 3)
# Metric for comparison will be accuracy for this project

metric <-  "Accuracy"
```



```{r echo=FALSE}
library(doParallel)
cl <- makePSOCKcluster(9)
registerDoParallel(cl)
```

```{r}

train_tree <- rpart(Diabetes_012 ~., data = train, method="class", control=rpart.control(cp=0, minsplit = 50, maxdepth = 5)) 

fancyRpartPlot(train_tree)
```

```{r}

#repeating a decision tree to exclude subjectivity from patient question on their general health
# Decision Tree
train2 <- subset(train,select=-c(GenHlth))


train_tree2 <- rpart(Diabetes_012 ~., data = train2, method="class", control=rpart.control(cp=0, minsplit = 50, maxdepth = 5)) 

fancyRpartPlot(train_tree2)
```



```{r echo=FALSE}
#Accuracy with GenHlth
trainPred = data.frame(predict(train_tree, train))
trainPred = as.data.frame(names(trainPred[apply(trainPred,1,which.max)]))
colnames(trainPred) = 'prediction'
trainPred$number = substr(trainPred$prediction, 2,2)
trainPred = train2 %>% bind_cols(trainPred) %>% select(Diabetes_012, number) %>% mutate(label=as.factor(Diabetes_012), number=as.factor(number))
confusionMatrix(trainPred$Diabetes_012, trainPred$number)
```

```{r}
#Accuracy without GenHlth
trainPred = data.frame(predict(train_tree2, train2))
trainPred = as.data.frame(names(trainPred[apply(trainPred,1,which.max)]))
colnames(trainPred) = 'prediction'
trainPred$number = substr(trainPred$prediction, 2,2)
trainPred = train2 %>% bind_cols(trainPred) %>% select(Diabetes_012, number) %>% mutate(label=as.factor(Diabetes_012), number=as.factor(number))
confusionMatrix(trainPred$Diabetes_012, trainPred$number)
```






```{r}
# Support Vector Machine (SVM)

svm.model <- train(Diabetes_012 ~ ., data = train2, method="svmRadial",metric=metric,trControl=control,
                   tuneLength = 5)
```

```{r echo=FALSE}
#Accuracy of SVMwithout GenHlth
SVMtrainPred = data.frame(predict(svm.model, train2))
colnames(SVMtrainPred) = 'prediction'
SVMtrainPred = train2 %>% bind_cols(SVMtrainPred) %>% select(Diabetes_012, prediction) %>% 
  mutate(label=as.factor(Diabetes_012), prediction)
confusionMatrix(SVMtrainPred$Diabetes_012, SVMtrainPred$prediction)
```

```{r echo=FALSE}
#Trying to improve accuracy by making classification as either health or flag for diabetes 
smp_size <- floor(0.66 * nrow(balanced))
balancedBinary <- balanced
balancedBinary$Diabetes_012[balancedBinary$Diabetes_012==2] <-  1
balancedBinary$Diabetes_012 <- as.factor(as.numeric(balancedBinary$Diabetes_012))

trainBinary_ind <- sample(seq_len(nrow(balanced)), size = smp_size)

trainBinary <- balancedBinary[trainBinary_ind, ]
testBinary <- balancedBinary[-trainBinary_ind, ]

# Creating a control with cross validation of 3
control <- trainControl(method ='cv',number = 3)
# Metric for comparison will be accuracy for this project

metric <-  "Accuracy"
```

```{r}
rf.modelbinary <- train(Diabetes_012 ~ ., data = trainBinary, method="rf", metric=metric, trControl=control,
                  tuneLength = 10)
knn.modelbinary <- train(Diabetes_012 ~ ., data = trainBinary, method="knn", metric=metric, trControl=control,
                   tuneLength = 5)
svm.modelBinary <- train(Diabetes_012 ~ ., data = trainBinary, method="svmRadial",metric=metric,trControl=control,
                   tuneLength = 5)

# summarize accuracy of models
```

```{r}
results <- resamples(list(Random_Forest=rf.modelbinary, knn=knn.modelbinary, SVM=svm.modelBinary))
dotplot(results)

```

```{r}
trainBinary <- subset(trainBinary,select=-c(GenHlth))
```

```{r}
train_treeBinary <- rpart(Diabetes_012 ~., data = trainBinary, method="class", control=rpart.control(cp=0, minsplit = 50, maxdepth = 4)) 
```

```{r}
fancyRpartPlot(train_treeBinary)
```
```{r}
#Accuracy without GenHlth
trainPred = data.frame(predict(train_treeBinary, trainBinary))
trainPred = as.data.frame(names(trainPred[apply(trainPred,1,which.max)]))
colnames(trainPred) = 'prediction'
trainPred$number = substr(trainPred$prediction, 2,2)
trainPred = trainBinary %>% bind_cols(trainPred) %>% select(Diabetes_012, number) %>% mutate(label=as.factor(Diabetes_012), number=as.factor(number))
confusionMatrix(trainPred$Diabetes_012, trainPred$number)
```
```{r}
#Accuracy without GenHlth
testPred = data.frame(predict(train_treeBinary, testBinary))
testPred = as.data.frame(names(testPred[apply(testPred,1,which.max)]))
colnames(testPred) = 'prediction'
testPred$number = substr(testPred$prediction, 2,2)
testPred = testBinary %>% bind_cols(testPred) %>% select(Diabetes_012, number) %>% mutate(label=as.factor(Diabetes_012), number=as.factor(number))
confusionMatrix(testPred$Diabetes_012, testPred$number)
```

```{r echo=FALSE}
svm.modelBinaryPRED = data.frame(predict(svm.modelBinary, testBinary))
colnames(svm.modelBinaryPRED) = 'prediction'
svm.modelBinaryPRED = testBinary %>% bind_cols(svm.modelBinaryPRED) %>% select(Diabetes_012, prediction) %>% 
  mutate(label=as.factor(Diabetes_012), prediction)
confusionMatrix(svm.modelBinaryPRED$Diabetes_012, svm.modelBinaryPRED$prediction)
``` 

